Bioverse: Origins of Life

### 2023-08-12 src/tex/ms.tex
rename P_abio, develop motivation

We may first consider the probability $\pdf(bio|H_i)$ of detecting a convincing biosignature on a planet, given a particular OoL hypothesis $H_i, \, i \in 1, 2$ is true.
This can be decomposed into the probability of abiogenesis $\pdf_\mathrm{abio}$, the fraction of life-hosting worlds that develop atmospheric biosignatures $\pdf_\mathrm{sig}$, the probability of the planet type required by the OoL hypothesis occurring in the surveyed sample $\pdf_\mathrm{\eta, i}$, and the probability of detecting the biosignature on this planet type with current technology $\pdf_\mathrm{det, i}$, yielding

\begin{align}
\label{eqn:pbio}
\pdf(bio|H_i) = \pdf_\mathrm{abio} \times \pdf_\mathrm{sig} \times \pdf_\mathrm{\eta, i} \times \pdf_\mathrm{det, i}.
\end{align}

However, what we are actually interested in is the probability of a OoL hypothesis being true, given a particular biosignature detection, $\pdf(H_i|bio)$.
To obtain this we can use Bayes' theorem, which yields
\begin{align}
\pdf(H_i|bio) = \frac{\pdf(bio|H_i) \pdf(H_i)}{\pdf(bio)}.
\end{align}
Here, $\pdf(H_i)$ is the prior probability of the OoL hypothesis $H_i$, and $\pdf(bio)$ is the prior probability of detecting a biosignature.
% Following Eq.~\ref{eqn:pbio}, we can sum over all possible scenarios to obtain
% \begin{align}
% \pdf(bio) = \sum_{i=1}^{N} \pdf_\mathrm{abio} \times \pdf_\mathrm{sig} \times \pdf_\mathrm{\eta, i} \times \pdf_\mathrm{det, i}.
% \end{align}

If the hypotheses $H_i$ are adjunct, i.e., their joint occurrence is impossible, one can show that
\begin{align}
\pdf(H_i|bio) = \frac{\pdf(bio|H_i) \pdf(H_i)}{\sum_{i=1}^{N} \pdf(bio|H_i) \pdf(H_i)}.
\end{align}
Then the parameters in Eq.~\ref{eqn:pbio} that are independent of the chosen hypothesis $H_i$ eliminate and we obtain
\begin{align}
\pdf(H_i|bio) = &\frac{\pdf_\mathrm{\eta, i} \pdf_\mathrm{det, i} \pdf(H_i)}{\sum_{i=1}^{N} \pdf_\mathrm{\eta, i} \pdf_\mathrm{det, i} \pdf(H_i)} \\
\overset{\pdf(H_i) = \pdf(H_j)}{ \underset{\forall i,j \in \{1, 2\}}{=}} &\frac{\pdf_\mathrm{\eta, i} \pdf_\mathrm{det, i}}{\sum_{i=1}^{N} \pdf_\mathrm{\eta, i} \pdf_\mathrm{det, i}},
\end{align}
where in the last step we made the implicit assumption that all OoL hypotheses are a priori equally probable.

If we take the ratio of these posteriors for our two independent hypotheses $H_1$ and $H_2$, we get the \textit{Bayes Factor}
\begin{align}
    \frac{\pdf(H_1|bio)}{\pdf(H_2|bio)} = \frac{\pdf_\mathrm{\eta, 1} \pdf_\mathrm{det, 1}}{\pdf_\mathrm{\eta, 2} \pdf_\mathrm{det, 2}},
\end{align}
which quantifies the evidence of the data arising from $H_1$ versus $H_2$.

% Putting everything together, we arrive at
% \begin{align}
% \label{eqn:posterior}
% \pdf(H_i|bio) = \frac{\pdf(H_i)}{\pdf(bio)} \times \pdf_\mathrm{abio} \times \pdf_\mathrm{sig} \times \pdf_\mathrm{\eta, i} \times \pdf_\mathrm{det, i}.
% \end{align}
% If we assume that all OoL hypotheses are a priori equally probable, we can treat $\frac{\pdf(H_i)}{\pdf(bio)}$ in Eq.~\ref{eqn:posterior} as a normalization constant.
In the following, we discuss the impact of the remaining variables $\pdf_\mathrm{\eta, i}$ and $\pdf_\mathrm{det, i}$ on the Bayes factor.

% \subsection{Probability of Abiogenesis $\pdf_\mathrm{abio}$}
% Following previous work~\citep{Spiegel2012,Chen2018,Kipping2021}, we may adopt a uniform rate model for abiogenesis, i.e., assume that OoL events occur at a uniform rate.
% This corresponds to a Poisson process with a rate parameter $lambda$, where we make the implicit assumption that abiogenesis occurs only via a single, unique mechanism, only once, and instantaneous.
% If this event occurs within a limited time window $t$, say, between planets form around a star and when it leaves the main sequence, we have
% \begin{align}
% \pdf_\mathrm{abio} = 1 - \exp(-\lambda t).
% \end{align}
% ~\\
% ...

% \subsection{biosignature fraction $\pdf_\mathrm{abio}$}
% Limiting ourselves to the search for \textit{life-as-we-know-it}, we may assume that there are atmospheric biomarkers present on a planet after the abiogenesis event and until global extinction.

>>>>

We may first consider the probability $\pdf(bio|H_i)$ of detecting a convincing biosignature on a planet, given a particular OoL hypothesis $H_i, \, i \in 1, 2$ is true.
This can be decomposed into the probability of abiogenesis in a particular environment $\pdf_\mathrm{env, i}$, the fraction of life-hosting worlds that develop atmospheric biosignatures $\pdf_\mathrm{sig}$, the probability of the planet type required by the OoL hypothesis occurring in the surveyed sample $\pdf_\mathrm{\eta, i}$, and the probability of detecting the biosignature on this planet type with current technology $\pdf_\mathrm{det, i}$, yielding

\begin{align}
\label{eqn:pbio}
\pdf(bio|H_i) = \pdf_\mathrm{env, i} \times \pdf_\mathrm{sig} \times \pdf_\mathrm{\eta, i} \times \pdf_\mathrm{det, i}.
\end{align}

However, what we are actually interested in is the probability of a OoL hypothesis being true, given a particular biosignature detection, $\pdf(H_i|bio)$.
To obtain this we can use Bayes' theorem, which yields
\begin{align}
\pdf(H_i|bio) = \frac{\pdf(bio|H_i) \pdf(H_i)}{\pdf(bio)}.
\end{align}
Here, $\pdf(H_i)$ is the prior probability of the OoL hypothesis $H_i$, and $\pdf(bio)$ is the prior probability of detecting a biosignature.
% Following Eq.~\ref{eqn:pbio}, we can sum over all possible scenarios to obtain
% \begin{align}
% \pdf(bio) = \sum_{i=1}^{N} \pdf_\mathrm{env, i} \times \pdf_\mathrm{sig} \times \pdf_\mathrm{\eta, i} \times \pdf_\mathrm{det, i}.
% \end{align}

If the hypotheses $H_i$ are adjunct, i.e., their joint occurrence is impossible, one can show that
\begin{align}
\pdf(H_i|bio) = \frac{\pdf(bio|H_i) \pdf(H_i)}{\sum_{i=1}^{N} \pdf(bio|H_i) \pdf(H_i)}.
\end{align}
Then the parameters in Eq.~\ref{eqn:pbio} that are independent of the chosen hypothesis $H_i$ eliminate and we obtain
\begin{align}
\pdf(H_i|bio) = &\frac{\pdf_\mathrm{env, i} \pdf_\mathrm{\eta, i} \pdf_\mathrm{det, i} \pdf(H_i)}{\sum_{i=1}^{N} \pdf_\mathrm{env, i} \pdf_\mathrm{\eta, i} \pdf_\mathrm{det, i} \pdf(H_i)} \\
\overset{\pdf(H_i) = \pdf(H_j)}{ \underset{\forall i,j \in \{1, 2\}}{=}} &\frac{\pdf_\mathrm{env, i} \pdf_\mathrm{\eta, i} \pdf_\mathrm{det, i}}{\sum_{i=1}^{N} \pdf_\mathrm{env, i} \pdf_\mathrm{\eta, i} \pdf_\mathrm{det, i}},
\end{align}
where in the last step we made the implicit assumption that all OoL hypotheses are a priori equally probable.

If we take the ratio of these posteriors for our two independent hypotheses $H_1$ and $H_2$, we get the \textit{Bayes Factor}
\begin{align}
    \frac{\pdf(H_1|bio)}{\pdf(H_2|bio)} = \frac{\pdf_\mathrm{env, 1} \pdf_\mathrm{\eta, 1} \pdf_\mathrm{det, 1}}{\pdf_\mathrm{env, 2} \pdf_\mathrm{\eta, 2} \pdf_\mathrm{det, 2}},
\end{align}
which quantifies the evidence of the data arising from $H_1$ versus $H_2$.

% Putting everything together, we arrive at
% \begin{align}
% \label{eqn:posterior}
% \pdf(H_i|bio) = \frac{\pdf(H_i)}{\pdf(bio)} \times \pdf_\mathrm{env, i} \times \pdf_\mathrm{sig} \times \pdf_\mathrm{\eta, i} \times \pdf_\mathrm{det, i}.
% \end{align}
% If we assume that all OoL hypotheses are a priori equally probable, we can treat $\frac{\pdf(H_i)}{\pdf(bio)}$ in Eq.~\ref{eqn:posterior} as a normalization constant.
In the following, we discuss the impact of the remaining variables $\pdf_\mathrm{env, i}$, $\pdf_\mathrm{\eta, i}$ and $\pdf_\mathrm{det, i}$ on the Bayes factor.

 \subsection{Required environment $\pdf_\mathrm{env, i}$}
% Following previous work~\citep{Spiegel2012,Chen2018,Kipping2021}, we may adopt a uniform rate model for abiogenesis, i.e., assume that OoL events occur at a uniform rate.
% This corresponds to a Poisson process with a rate parameter $lambda$, where we make the implicit assumption that abiogenesis occurs only via a single, unique mechanism, only once, and instantaneous.
% If this event occurs within a limited time window $t$, say, between planets form around a star and when it leaves the main sequence, we have
% \begin{align}
% \pdf_\mathrm{env, i} = 1 - \exp(-\lambda t).
% \end{align}
% ~\\
\todo[inline]{The planet should be in the liquid-water HZ}
...
Let us assume that $H_1$ only requires a minimum bulk density $\rho_1$, such that $\pdf_\mathrm{env, 1} = 1$ if $\rho > \rho_1$ and zero otherwise.
On the other hand, $H_2$ requires exposed land and thus a small water mass fraction.
We may implement this in the same way as above but with a minimum bulk density $\rho_2 > \rho_1$.
Furthermore, there is a requirement that the tidal locking timescale may not be so small that the planet is likely tidally locked at observation.
This translates to imposing a minimum semimajor axis $a_2$.
...
...

% \subsection{biosignature fraction $\pdf_\mathrm{env, i}$}
% Limiting ourselves to the search for \textit{life-as-we-know-it}, we may assume that there are atmospheric biomarkers present on a planet after the abiogenesis event and until global extinction.
